---
title: "grad project"
output: html_notebook
author: "Laura Cech"
date: "Dec. 4, 2023"

---

I looked at the data 
https://opendata.maryland.gov/Business-and-Economy/Baltimore-County-Real-Property-Assessments-Hidden-/jpfc-qkxp
based on SDAT (State Department of Assessments and Taxation) and MDP (Maryland Department of Planning) records.
It is updated monthly.
There are 220 columns and more than 11,000 records.

There are dates included, but the data is organized by Sales 1, Sales 2 and Sales 3, showing the most recent transactions, the transactions before that and the one before that.

The most impressive part of my analysis is that when I cleaned "Ryland," I didn't erroneously get "Maryland."

You should explore the data to find at least 3 potential story ideas, and conduct enough analysis to describe the three best ideas you have in pitches. To do this, you should join it with one other source of information - that could be Census data or some other type of information that is related to the data you are working with.

Pitch 1: Baltimore County is known as a suburb but a surprising number of properties are sold without access to public services such as water and sewer. Some of the land will remained undeveloped, but some will be connected to public water and sewer. What's involved in connecting to the public works grid? Is there a cost to all taxpayers? Are there any houses left with just outhouses? 
See data visualization: https://datawrapper.dwcdn.net/Zl4fY/1/

Pitch 2: While most property transactions are by individuals and corporations, there are some companies and agencies who make more than few purchases. Any one of them could be a potential story. For example, are the Secretary of Housing transactions really just mortgages? Are they clustered in a certain area? 
We could also look at the campaign contributions made by developers.

Of the most recent transactions, top buyers were: NVR (1462), RYLAND (1455), SECRETARY OF HOUSING (1,300)
See visualization: https://datawrapper.dwcdn.net/Y7QBW/1/

Pitch 3: George Washington didn't sleep in any of the townhouses in Dundalk, or stop at an industrial space in Middle River. But they are among four properties listed as having significant historical influence in property records. An office building in Lutherville-Timonium was built in 1924, but if the building's age qualifies it as "historic" then hundreds, maybe thousands more, should be on the list. Why doesn't the historical influence data doesn't seem accurate?

Note:
My waterfront analysis didn't really turn up anything interesting. See the by-the-city visualization: https://datawrapper.dwcdn.net/eimgT/1/ But if I analyzed by date, I might have been able to show how Baltimore County's waterfront is changing.


#intial set up, load libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)
library(readxl)
library(datasets)
library(dplyr)
library(stringr)
```

```{r}
property_all <- read_csv("data/Baltimore_County_Real_Property_Assessments.csv")
```
# initial clean; refining of owners on line 352

```{r}
property_all <-property_all|>
  clean_names()
```
```{r}
property_filtered <- property_all %>% 
  select(mappable_latitude_and_longitude, mdp_street_address_mdp_field_address, mdp_street_address_city_mdp_field_city, mdp_street_address_zip_code_mdp_field_zipcode, land_use_code_mdp_field_lu_desclu_sdat_field_number_50, property_factors_utilities_water_mdp_field_pfuw_sdat_field_number_63, property_factors_utilities_sewer_mdp_field_pfus_sdat_field_number_64, property_factors_location_waterfront_mdp_field_pflw_sdat_field_number_65, property_factors_street_unpaved_mdp_field_pfsu_sdat_field_number_68, property_factors_influence_historical_mdp_field_pfih_sdat_field_number_70, sales_segment_1_grantor_name_mdp_field_grntnam1_sdat_field_number_80, sales_segment_3_how_conveyed_ind_sdat_field_number_127, 
sales_segment_2_grantor_name_sdat_field_number_100, 
sales_segment_3_grantor_name_sdat_field_number_120, additional_c_a_m_a_data_building_style_code_and_description_mdp_field_strustyl_descstyl_sdat_field_number_264  )
```

#Note on the columns I pulled: I did not look at prices at this juncture and only looked at the type of the most recent sale. There are too many things to examine and the file is huge!

```{r}
#rename columns
property_renamed <-property_filtered |>
  rename(map_coordinates = mappable_latitude_and_longitude) |>
  rename(street = mdp_street_address_mdp_field_address) |>
  rename(city = mdp_street_address_city_mdp_field_city) |>
  rename(zip = mdp_street_address_zip_code_mdp_field_zipcode) |>
  rename(land_code = land_use_code_mdp_field_lu_desclu_sdat_field_number_50) |>
  rename(water = property_factors_utilities_water_mdp_field_pfuw_sdat_field_number_63) |>
  rename(sewer = property_factors_utilities_sewer_mdp_field_pfus_sdat_field_number_64) |>
  rename(waterfront = property_factors_location_waterfront_mdp_field_pflw_sdat_field_number_65) |>
  rename(unpaved = property_factors_street_unpaved_mdp_field_pfsu_sdat_field_number_68) |>
  rename(historical = property_factors_influence_historical_mdp_field_pfih_sdat_field_number_70) |>
  rename(sales1 = sales_segment_1_grantor_name_mdp_field_grntnam1_sdat_field_number_80) |>
  rename(sales2 = sales_segment_2_grantor_name_sdat_field_number_100) |>
  rename(sales3 = sales_segment_3_grantor_name_sdat_field_number_120) |>
  rename(conveyed3 = sales_segment_3_how_conveyed_ind_sdat_field_number_127) |>
  rename(bldg = additional_c_a_m_a_data_building_style_code_and_description_mdp_field_strustyl_descstyl_sdat_field_number_264)
  
```

# What does the data look like by city?
I can see that some places will show up as "Baltimore" because it's interchangable in some places like Catonsville.

```{r}
#group by cities
by_city <- property_renamed |>
   group_by(city) |>
  summarise(count = n()) |>
  arrange(desc(count))
```

# What percentage of this data involves properites with historical influence?
I'm surprised only 4 are in this category, but it made it easy to research the properties to see if there's a story there. It appears those listed as historical influence are typos. They don't seem to be historic at all.


```{r}
#see historical properties
historical_count <- property_renamed |> 
  filter(!str_detect(historical, "No Historical Influence \\(0\\)")) 

print(historical_count)
```

# Where is Baltimore County's waterfront?

Only about 5,000 (less than 2 percent) of the 305,205 records are classified as waterfront.
Obviously, the majority are in Eastern Baltimore County (Dundalk, Middle River, Essex).

See the by-the-city visualization: https://datawrapper.dwcdn.net/eimgT/1/


```{r}
waterfront_total <- property_renamed |>
  filter(!str_detect(waterfront, "None \\(0\\)")) |>
  nrow()

#total records
total_records <- nrow(property_renamed )

#get percentage 
percentage_waterfront <- (waterfront_total / total_records) * 100

print(percentage_waterfront)
```

```{r}
waterfront_counts_by_city <- property_renamed |>
  filter(!str_detect(waterfront, "None \\(0\\)")) |>
  group_by(city, land_code) |>
  summarise(count = n()) |>
  arrange(desc(city))


print(waterfront_counts_by_city)

# Export csv
write.csv(waterfront_counts_by_city, "waterfront_counts_by_city.csv", row.names = FALSE)

```
```{r}
#see land code percentages by city

waterfront_counts_by_city_landcode <- property_renamed |>
  filter(!str_detect(waterfront, "None \\(0\\)")) |>
  group_by(city, land_code) |>
  summarise(count = n()) |>
  arrange(desc(city))

total_counts_by_city <- waterfront_counts_by_city_landcode |>
  group_by(city) |>
  summarise(total_count = sum(count))

# calculate percentages
waterfront_percentages <- waterfront_counts_by_city_landcode |>
  left_join(total_counts_by_city, by = "city") |>
  mutate(percentage = (count / total_count) * 100)

# View the result
print(waterfront_percentages)

#export
write.csv(waterfront_percentages, "waterfront_percentages.csv", row.names = FALSE)
```
```{r}
#asked chat gpt for code to show zip codes too
waterfront_counts_by_city2 <- property_renamed |>
  filter(!str_detect(waterfront, "None \\(0\\)")) |>
  group_by(city, land_code) |>
  summarise(
    count = n(),
    zip = paste(unique(zip), collapse = ", "),
    .groups = 'drop'
  ) |>
  arrange(desc(city))

glimpse(waterfront_counts_by_city2)
```
```{r}
#export
write.csv(waterfront_counts_by_city2, "waterfront_counts_by_city2.csv", row.names = FALSE)
```

See new visualization:

# What are landcodes of waterfront properties?
I thought it would be cool to show pies for each city in datawrapper, but it didn't work even after I eliminated the cities with only one landcode. I think maybe the reason is that most of the sales types (90percent-ish) are residential.

```{r}
waterfront_landcodes <- waterfront_percentages |>
  group_by(city) |>
  filter(n_distinct(land_code) > 1)

# View the resulting DataFrame
print(waterfront_landcodes)

#export
write.csv(waterfront_landcodes, "waterfront_landcodes.csv", row.names = FALSE)
```

# What can landcode data tell us in general? 
81 percent of the transactions are residential. 

Here's a visualization of landcodes that aren't residential: https://datawrapper.dwcdn.net/49Hc3/1/


```{r}

# count land code
land_code_counts <- property_renamed |>
  count(land_code)

# total 
total_records <- nrow(property_renamed)

# percentages
land_code_percentages <- land_code_counts |>
  mutate(percentage = n / total_records * 100)

# view
print(land_code_percentages)

#export
write.csv(land_code_percentages, "land_code_percentages.csv", row.names = FALSE)

```
```{r}
#see non-residential only

non_residential <- property_renamed |>
  filter(land_code != "Residential (R)" , land_code != "Residential Condominium (U)")

# count
land_code_counts_non <- non_residential |>
  count(land_code)

# number of non-residential records
total_non_residential <- nrow(non_residential)

# percentages
land_code_percentages_non <- land_code_counts_non |>
  mutate(percentage = n / total_non_residential * 100)

# view the land code percentages
print(land_code_percentages_non)

#export
write.csv(land_code_percentages_non, "land_code_percentages_non.csv", row.names = FALSE)

```

# Do some residential properties not have public services like sewer, water and paved roads?
245,798 have unpaved roads, 68,050 have no public sewer, and 245,798 don't have public water.
It would be nice to have a comparison with other counties for this, but my computer is puny and can't handle the data.

```{r}
#roads
count_not_unpaved <- property_renamed |>
  filter(unpaved != "No Paved Street (0)", land_code == "Residential (R)") |>
  nrow()

print(count_not_unpaved)

not_unpaved <- property_renamed |>
  filter(unpaved != "No Paved Street (0)", land_code == "Residential (R)") 
```

```{r}
#sewer
count_no_sewer <- property_renamed |>
  filter(sewer != "Public Sewer (1)", land_code == "Residential (R)") |>
  nrow()

print(count_no_sewer)

no_sewer <- property_renamed |>
  filter(sewer != "Public Sewer (1)", land_code == "Residential (R)")
```

```{r}
#water
count_no_water <- property_renamed |>
  filter(sewer != "Public Water (1)", land_code == "Residential (R)") |>
  nrow()

print(count_no_water)

no_water <- property_renamed |>
  filter(sewer != "Public Water (1)", land_code == "Residential (R)")

```

```{r}
no_service <- bind_rows(no_water, no_sewer, not_unpaved)

glimpse(no_service)
```

```{r}
no_service_filtered <- no_service |>
group_by(zip) |>
  summarise(count = n()) |>
  arrange(desc(count))
```


```{r}
#export
write.csv(no_service_filtered, "no_service_filtered.csv", row.names = FALSE)
```

See visualization: https://datawrapper.dwcdn.net/Zl4fY/1/

Owner trends
#How are properties conveyed? (arms length, etc.)
When I created the visualization https://datawrapper.dwcdn.net/GRILY/1/, I realized it would be more valuable to see the "Unknown" and "NA" together and all the non-arms length transactions together: https://datawrapper.dwcdn.net/WT3sk/1/

Next I examined the most transactions first (sales3) to see what companies/ government agencies emerged as large land owners.

```{r}
conveyed <- property_renamed |>
group_by(conveyed3) |>
 summarise(count = n())

glimpse(conveyed)
```
```{r}
#chat gpt help in how to mutate "No Data" and "NA" into an "Unkown" result and all the arms-length together.


conveyed1 <- conveyed %>%
  mutate(
    conveyed3 = case_when(
      conveyed3 %in% c("No Data", "na") ~ "Combined No Data/na",
      grepl("Private Non-Arms-Length", conveyed3, ignore.case = TRUE) ~ "Private Non-Arms-Length",
      grepl("Private Arms-Length", conveyed3, ignore.case = TRUE) ~ "Private Arms-Length",
      TRUE ~ "Unknown"
    )
  ) %>%
  group_by(conveyed3) %>%
  summarize(count = sum(count))

print(conveyed1)


#export
write.csv(conveyed1, "conveyed1.csv", row.names = FALSE)
```



```{r}

# count properties per owner
owners_count <- property_renamed |>
  count(sales3) |>
  filter(n > 1) |>
  arrange(desc(n))

# view
print(owners_count)

```


```{r}
#checking dupes (note: I also tried n_gram_merge() but it made the data worse. I also tried open refine and it didn't find clusters.)
owners_count_dupes <-owners_count |>
   get_dupes() 

glimpse(owners_count_dupes)

#clean names
  cleaned_owners_count <- owners_count |>
  mutate(owners_clean=key_collision_merge(sales3)) |>
  select(sales3, owners_clean, everything())

cleaned_owners_count
```

```{r}
#export for open refine
write.csv(cleaned_owners_count, "cleaned_owners_count.csv", row.names = FALSE)
```



```{r}
#grouping nvr
nvr <- cleaned_owners_count |>
  filter(str_detect(owners_clean, "NVR")) 
         
glimpse(nvr)
```
```{r}
#standardize nvr
cleaned_owners <- cleaned_owners_count |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, "NVR"), "NVR", owners_clean))

```

```{r}
#standardize Ryland without getting "Maryland" in the mix with the help of chat gpt
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, "\\bRyland\\b"), "RYLAND", owners_clean))

```

```{r}
#standardize Veterans Affairs
cleaned_owners <-cleaned_owners |>
  mutate(owners_clean = ifelse(
    str_detect(owners_clean, "ADMINISTRATOR OF VETERANS AFFAIRS|SECRETARY OF VETERANS AFFAIRS|THE SECRETARY OF VETERANS"), 
    "VETERANS AFFAIRS", 
    owners_clean))

```

```{r}
#standardize TITO INC and TITO REALTY LLC
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, "TITO INC|TITO REALTY LLC"), "TITO", owners_clean))

```

```{r}
#standardize SECRETARY OF HOUSING
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
     "SECRETARY OF HOUSING( AND| &)?( URBAN( DEV)?)?|THE SECRETARY OF HOUSING( AND| &)?( URBAN( DEV)?)?"), 
                               "SECRETARY OF HOUSING", 
                               owners_clean))

```


```{r}
#standardize HIGHPOINT BUSINESS TRUST and HIGHPOINTE BUSINESS TRUST
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, "HIGHPOINTE? BUSINESS TRUST"), 
                               "Highpoint Business Trust", 
                               owners_clean))
```


```{r}
#standardize HSBC BANK USA, HSBC BANK USA N A, HSBC BANK USA N A TRUSTEE, HSBC BANK USA N A,TRUSTEE, HSBC BANK USA NA TRUSTEE, HSBC BANK USA NATIONAL, HSBC BANK USA NATIONAL ASSOCIATION, HSBC BANK USA TRUSTEE
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
                               "HSBC BANK USA( N A| NATIONAL ASSOCIATION|,? TRUSTEE| NA TRUSTEE| NATIONAL)?"), 
                               "HSBC Bank USA", 
                               owners_clean))
```

```{r}
#standardize JOHNNYCAKE DEVEL OPMENT CORP, JOHNNYCAKE DEVEL OPMENT CORPORATIO, JOHNNYCAKE DEVELOPMENT CORP, JOHNNYCAKE DEVELOPMENT CORPORATION
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
                               "JOHNNYCAKE DEVEL\\s?OPMENT\\s?(CORP(ORATION)?)?"), 
                               "Johnnycake Development Corp", 
                               owners_clean))

```


```{r}
#standardize K HOVNANIAN AT EDEN TERRACE, K HOVNANIAN HOMES, K HOVNANIAN HOMES OF, K HOVNANIAN HOMES OF MARYLAND, K HOVNANIAN HOMES OF MARYLAND I LL, K HOVNANIAN HOMES OF MARYLAND L L, K HOVNANIAN HOMES OF MARYLAND LLC
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
                               "K HOVNANIAN (AT EDEN TERRACE|HOMES( OF( MARYLAND( I LL| L L| LLC)?)?)?)"), 
                               "K Hovnanian", 
                               owners_clean))

```

```{r}
#standardize WELLS FARGO BANK, WELLS FARGO BANK MINNESOTA NA, WELLS FARGO BANK MINNESOTA,TRUSTEE, WELLS FARGO BANK N A, WELLS FARGO BANK N A TRUSTEE, WELLS FARGO BANK N A TRUSTEE, WELLS FARGO BANK NA, WELLS FARGO BANK NA TRUSTEE, WELLS FARGO BANK NATIONAL, WELLS FARGO BANK NATIONAL ASSOCIAT, WELLS FARGO BANK TRUSTEE, WELLS FARGO BANK, NA, WELLS FARGO BANK NA, WELLS FARGO BANK,NA,TRUSTEE, WELLS FARGO BANK NA TRUSTEE, WELLS FARGO FINANCIAL MARYLAND, WELLS FARGO FINANCIAL MARYLAND INC
cleaned_owners <- cleaned_owners |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
                               "WELLS FARGO BANK( MINNESOTA( NA)?,?| N A| NA| NATIONAL( ASSOCIAT)?|,? NA|,?TRUSTEE| NA TRUSTEE)?|WELLS FARGO FINANCIAL MARYLAND( INC)?"), 
                               "WELLS FARGO",  
                               owners_clean))


```

```{r}
#standardize BALTIMORE COUNTY, BALTIMORE COUNTY  MARYLAND, BALTIMORE COUNTY  MD, BALTIMORE COUNTY MARYLAND, BALTIMORE COUNTY MD.,BALTIMORE COUNTY, MARYLAND

cleaned_owners <- cleaned_owners |>
   mutate(owners_clean = ifelse(str_detect(owners_clean, 
                                         "BALTIMORE COUNTY(,? MARYLAND|,? MD|,? MD\\.|)?$"), 
                               "Baltimore County", 
                               owners_clean))

```


```{r}
owners_most <- cleaned_owners |>
  group_by(owners_clean) |>
  summarise(total_n = sum(n, na.rm = TRUE)) |>
  arrange(desc(total_n))

```


```{r}
glimpse(owners_most)
```

```{r}
#standardize more to fix Ryland but not get Maryland
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
          "\\bRYLAND GROUP\\b|\\bRYLAND GROUP INC ORPORATED\\b|\\bRYLAND GROUP INC THE\\b"), 
          "RYLAND", 
          owners_clean))

```

```{r}
#standardize more to fix Pulte

owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
         "PULTE HOME CORP|PULTE HOME CORPO RATION|PULTE LAND OF MA RYLAND CORPORATIO"), 
         "PULTE", 
          owners_clean))

```


```{r}
#standardize to fix Baltco again

owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
                                          "(?i)\\bBALTIMORE COUNTY\\b\\s*(MD|MARYLAND)?$"), 
                               "BALTIMORE COUNTY", 
                               owners_clean))


```

```{r}
#standardize Ashland
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
           "ASHLAND JOINT VENTURE|ASHLEY CUSTOM HOMES INC|ASHLEY GROUP LIM ITED|ASHLEY GROUP LTD"), 
           "ASHLEY GROUP", 
                               owners_clean))

```

```{r}
#standardize Associated
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
          "ASSOCIATED JEWIS H CHARITIES &|ASSOCIATED JEWISH CHARITIES"), 
          "ASSOCIATED JEWISH", 
           owners_clean))

```


```{r}
#standardize Bancroft
owners_most <- owners_most |>
   mutate(owners_clean = ifelse(str_detect(owners_clean, 
     "BANCROFT HOMES|BANCROFT HOMES O F|BANCROFT HOMES O F HUNT VALLEY INC|BANCROFT HOMES O F JONES VALLEY IN|BANCROFT PROPERT IES|BANCROFT PROPERTIES"), 
                               "BANCROFT HOMES", 
                               owners_clean))
```



```{r}
#standardize Bank of America
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
         "BANK OF AMERICA( N A| N\\.A\\.| NA| NATIONAL| NATIONAL ASSOCIATI|, N A)?"), 
         "BANK OF AMERICA", 
          owners_clean))
```

```{r}
#standardize Bank of New York
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
         "BANK OF AMERICA( N A| N\\.A\\.| NA| NATIONAL| NATIONAL ASSOCIATI|, N A)?"), 
         "BANK OF AMERICA", 
                               owners_clean))
```

```{r}
#standardize Beazer
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
          "BEAZER HOME CORP|BEAZER HOMES CORP|BEAZER HOMES CORPORATION|BEAZER HOMES LLC"), 
          "BEAZER HOMES", 
          owners_clean))
```

```{r}
#standardize DEUTSCHE BANK
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, "DEUTSCHE"), 
                               "DEUTSCHE", 
                               owners_clean))
  
```

```{r}
#standardize federal home
owners_most <- owners_most |>
   mutate(owners_clean = ifelse(str_detect(owners_clean, 
            "FEDERAL HOME( LOAN( MORTGAGE( CORP(ORA)?)?)?)?\\b"), 
             "FEDERAL HOME", 
               owners_clean))
  
```

```{r}
#standardize federal national
owners_most <- owners_most |>
  mutate(owners_clean = ifelse(str_detect(owners_clean, 
 "FEDERAL NATIONAL( MORGAGE ASSOCIATI| MORTAGE ASSOCIATI| MORTG ASSN| MORTGAGE| MORTGAGE ASSN| MORTGAGE ASSOC| MORTGAGE ASSOCIAT)?"), 
  "FEDERAL NATIONAL", 
  owners_clean))
 
```



```{r}
#re-run count
owners_most_again <- owners_most |>
  group_by(owners_clean) |>
  summarise(total_sum = sum(total_n, na.rm = TRUE)) %>%
  arrange(desc(total_sum))

glimpse(owners_most_again)

#remove na
owners_most_again  <- owners_most_again |>
  filter(!is.na(owners_clean))

#only show 200 and more
top_owners  <- owners_most_again |>
   filter(total_sum > 200)
  
#export 
write.csv(top_owners, "top_owners.csv", row.names = FALSE)
```

```{r}


```

