---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv")
problems(payments)
# Load in the Maryland grants and loan data
grants <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")
problems(grants)

```

```{r}
# cleaning function
cleaned_payments <- payments |>
  clean_names()

# display the cleaned dataset
cleaned_payments
```


```{r}
# cleaning function
cleaned_grants <- grants |>
  clean_names()

# display the cleaned dataset
cleaned_grants
```


## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system. 

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? 

A1: 60

```{r}
pg_schools <- cleaned_payments |>
filter(str_detect(vendor_name, "SCHOOL" )) |>
filter(str_detect(vendor_name, "PRINCE" )) 
 

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2021	1,470,133,533	
There appear to be missing years.

```{r}
pg_schools |>
   group_by(fiscal_year) |>
  summarize(total = sum(amount)) |>
  arrange(desc(total))


```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. 

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.
2008	DEPARTMENT OF NATURAL RESOURCES	CANON FINANCIAL SERVICES INC	
2008	DEPARTMENT OF NATURAL RESOURCES	CANON FINANCIAL SERVICES INC	
2010	BOARDS COMMISSIONS & OFFICES	HASLER FINANCIAL SERVICES	
2010	BOARDS COMMISSIONS & OFFICES	HASLER FINANCIAL SERVICES	
2019	DEPARTMENT OF JUVENILE SERVICES-HQ	LABORATORY CORPORATION OF AMERICA	
2019	DEPARTMENT OF JUVENILE SERVICES-HQ	LABORATORY CORPORATION OF AMERICA	

Use distinct() to remove duplicate rows in tidyverse



```{r}
cleaned_payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND" )) |>
    arrange(desc(vendor_name)) 

cleaned_payments |>
  distinct()

  cleaned_payments |>
  group_by(vendor_name) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
  

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research spending has decreased since 2009 when it peaked at 14,643,721.	

```{r}
stemcell <-cleaned_grants |>
   filter(str_detect(description, "Stem Cell Research Grant") | str_detect(description, "STEM CELL RESEARCH GRANT")) 
  
```



```{r}
stemcell |>
  
  group_by(fiscal_year) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 18 grants totaling 7,401,905.35. Nine grants are going to the University of Maryland. I used the "distinct" function to avoid duplicates. I could also have capitalized all UMD and University of Maryland grantees. If it doesn't matter where in the university the grants are doing, I could mutate the data so that all variations of UMD and University Maryland became a single UMD data. I'm curious about whether the tech companies (all the non-UMD grants) are somehow tied to UMD. I'm also curious if there are any grants to the UMD Foundation, which does not use the UMD campus zipcode.

```{r}
umdgrants <- cleaned_grants |>
  filter(str_detect(zip_code, "20742" )) |>
  distinct() |>
  
   group_by(grantee) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount)) 
 
 


```


```{r}
umdgrants <- cleaned_grants |> 
  filter(str_detect(as.character(zip_code), "20742")) |> 
  distinct() |> 
  group_by(grantee) |> 
  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = "drop") |> 
  arrange(desc(total_amount))

# Calculating and binding the grand total row
grand_total <- umdgrants %>%
  summarise(grantee = "Total", total_amount = sum(total_amount, na.rm = TRUE))

umdgrants_total <- bind_rows(umdgrants, grand_total)

# Display the result
print(umdgrants_total)







```

